---
title: "Protocol"
output: 
  md_document:
    variant: markdown_github
---

<!-- badges: start -->

[![GPLv3 license](https://img.shields.io/badge/License-GPLv3-blue.svg)](http://perso.crans.org/besson/LICENSE.html) [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-experimental-red.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

<!-- badges: end -->

# Introduction

# Preliminary steps

## Verify sequencig quality

To verify the quality of the sequencing results, there are several tools available.

We recommend using either FASTQC (Andrews, 2010) or MultiQC (Ewels, 2016).

-   FASTQC: <https://www.bioinformatics.babraham.ac.uk/projects/fastqc/>
-   MultiQC: <https://seqera.io/multiqc/>

## Pre-processing of FASTQ files

If the FASTQ files include adaptor sequences and/or primers, it is possible to remove them using cutadapt, for example.

-   cutadapt: <https://cutadapt.readthedocs.io/en/stable/>

Primer removal is also possible in the DADA2 section of code, presented below. However, if you remove the primers with cutadapt, then you must not cut them again in DADA2.

# Obtain unique sequences using DADA2

DADA2 is an R package used to assign amplicon sequence variants (ASVs) from FASTQ files (Callahan et al., 2016).

## Data preparation
The first few steps will ensure that DADA2 knows where the FASTQ files are stored and what they refer to. Note that you will need to change the path according to your own files.

```{r, eval = FALSE}
path <- "./ATLANTIDA_eDNA" # CHANGE ME to the directory containing the fastq files after unzipping.
# verify files in path
list.files(path)

# Forward and reverse fastq file names have format: SAMPLENAME_R1_001.fastq and SAMPLENAME_R2_001.fastq
# CHANGE according to your file names
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnFs
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
fnRs

# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.namesF <- sapply(strsplit(basename(fnFs), "[.]"), `[`, 1)
sample.namesF
sample.namesR <- sapply(strsplit(basename(fnRs), "[.]"), `[`, 1)
sample.namesR

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.namesF, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.namesR, "_R_filt.fastq.gz"))
names(filtFs) <- sample.namesF
names(filtRs) <- sample.namesR
```

## Quality profiles

Inspect quality of sequencing.

```{r, eval = FALSE}
# Example for 5 samples
# Quality of forward reads
plotQualityProfile(fnFs[1:5])
# Quality of reverse reads
plotQualityProfile(fnRs[1:5])

# To inspect many samples at once
plotQualityProfile(fnFs, aggregate = TRUE)
plotQualityProfile(fnRs, aggregate = TRUE)
```

## Filter and trim reads

Based on quality profiles, decide the trimming parameters. To remve nucleotides in the end of the reads, set the parameter *truncLen*, where the first value corresponds to the trimming of the forward reads and the second is for the reverse reads.

If the primers are present in your samples and you are sure that they are right at the beggining of the sequence, then you can use *trimLeft* to remove them.

All other parameters are set to default.

**Note:** If you are using a Windows OS, set multithread to FALSE.

```{r, eval = FALSE}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(240,210), 
                     trimLeft = c(nchar("AGACGAGAAGACCCTATG"), ##remove primers
                                  nchar("GGATTGCGCTGTTATCCC")), 
                     maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE, 
                     compress=TRUE, multithread=FALSE) # On Windows set multithread=FALSE
```

## Learn error rates

To distinguish true sequence variations from sequencing errors, DADA2 calculates the probability of finding an error, given the error distribution. So, the next step is to learn the error rates:

**Note:** This step might take a while. Again, set multithread to FALSE, if using Windows OS.

```{r, eval = FALSE}
#Learn the Error Rates
errF <- learnErrors(filtFs, multithread=FALSE)
errR <- learnErrors(filtRs, multithread=FALSE)
```

After learning the error rates, it is possible to do a sanity check on the model:

```{r, eval = FALSE}
plotErrors(errF, nominalQ=TRUE)
plotErrors(errR, nominalQ=TRUE)
```

## Obtain ASVs

Based on error rates model, DADA will identify unique sequences:

```{r, eval = FALSE}
# Identify unique sequences
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

Next, DADA2 will merge the forward and reverse reads. If after this step you lost a significant ammount of reads, check the trimming parameters.

```{r, eval = FALSE}
#Merge paired reads
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
```

Construct an abundance table:

```{r, eval = FALSE}
#Construct sequence table
seqtab <- makeSequenceTable(mergers)
```

Remove reads outisde the target length:

```{r, eval = FALSE}
# Remove non-target-length sequences from your sequence table
seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 200:300]
```

## Remove chimeric sequences

To remove chimeric sequences:

```{r, eval = FALSE}
#Remove chimeras
seqtab.nochim <- removeBimeraDenovo(seqtab2, method="consensus", multithread=TRUE, verbose=TRUE)
```

## Summary track reads

Then we can track the number of reads after each step:

```{r, eval = FALSE}
#Track reads through the pipeline
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))

# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
```

## Save ASV table

At this stage, you can save the ASV table for later use:

```{r, eval = FALSE}
#Create .csv file
write.table(seqtab.nochim, file='ASV_table.tsv', quote=FALSE, sep='\t', col.names = NA)
```

## Export reads to a FASTA file

Generally, it is useful to have the final unique sequences in a FASTA file. We are going to use this file later for BLASTN.

In this step, it is important to keep track of the position of the ASVs, so that we can connect the taxonomy obtained with NCBI and the abundance table.

```{r, eval = FALSE}
# make FASTA file
library(seqinr)

# load ASV table
ASV_table <- read.table("./ASV_table.tsv") ## this is the table we saved in the previous session

# make data frame with unique ASVs ID and Sequence
ASVs.df <- ASV_table %>% 
    colnames() %>% 
    as.data.frame() %>% 
    rename(Sequence = ".") %>% 
    distinct() %>% 
  mutate(ASV = paste0("ASV_", row_number(.)))

# Make FASTA file 
write.fasta(sequences = as.list(ASVs.df$Sequence), 
            names = ASVs.df$ASV, 
            "./ASV.fasta",
            as.string = TRUE)
```

# Assign taxonomy

To assign taxonomy, we follow these steps:

1.  BLASTN against the mitochondrial 16S database from NCBI (link).
2.  Filter the best hits based on multiple parameters (see below).
3.  Solve ties within genus and family level (for *Delphinidae* sp.)

## Run BLASTN against NCBI

To run blastn (Camacha et al., 2009; Altschul et al., 1990) against the nucleotide database of NCBI (Benson et al., 2013), we use the **BLAST Command Line Tool**.

Please see installation instructions at: <https://www.ncbi.nlm.nih.gov/books/NBK569861/>

Blast parameters: 
 -   Minimum percentage identity: 99.0%
 -   Maximum evalue: 10⁻⁵
 -   Minimum query cover: 80%

```{r, engine = 'bash', eval = FALSE}
blastn -db nt -query ASV_atlantida.fasta -out blast_results_atlantida -outfmt "6 delim=, qacc qlen sseqid sacc slen evalue bitscore score length pident nident mismatch positive gaps staxid ssciname sblastname scomnames skingdoms" -evalue 1e-05 -perc_identity 99 -qcov_hsp_perc 80 -remote
```

This command returns a table named blast_results (you can change the name as needed). The parameter *-outfmt* determines the format and variables present in the table.

The parameter *-remote* runs the code in the NCBI dedicated server, which means that the time it takes to run your samples might vary (it may take more than 1 hour).

**Note:** change the file paths as needed.

## Assign taxonomy based on best hits

The raw blast results include all the hits. Therefore, we need to apply multiple filters to obtain the best hits. To do so, we go back to R.

Start by loading the blast results into your R session:

```{r, eval = FALSE}
# load packages
library(dplyr)
library(tidyr)
library(ulrb)
library(stringr)
library(purrr)

# load blast results
all_hits <- read.csv("./blast_results", header = FALSE, # change file path as needed
                     col.names = c("Query accession", "Query sequence length",
                                   "Subject seq-id",	"Subject accession",
                                   "Subject sequence length",	"evalue", "Bit Score",
                                   "Raw Score",	"Alignment length", "Percentage of identical matches",
                                   "Number of identical matches",	"Number of mismatches",
                                   "Number of positive scoring matches", "Total number of gaps",
                                   "Taxonomy ID", "Scientific name",	"Subject blast name", 
                                   "Subject common name"))
```

Based on domain knowledge of the area under study, we can assume that some hits are probably wrong. Therefore, to avoid False positive detections, we issue a ban list on some species.

Feel free to edit the list as needed for your experimental setting, by editing the file **ban_list**.

```{r, eval = FALSE}
# ban list
ban_list <- read.table("./refs/ban_list.txt", header = FALSE) %>% 
  rename(Genus = V1,
         Species = V2) %>% 
  mutate(binomial_name = paste(Genus, Species)) %>% 
  pull(binomial_name)
```

Additionally, we also need to ensure that we are not obtaining matches from non-16S genes. To do so, we filter all accessions based on a reference file with all possible target gene accessions.

```{r, eval = FALSE}
# target genes
target_genes <- read.table("refs/gene_16_list", header = FALSE) ## last accessed 23 May 2025
# some data cleaning
target_genes <- target_genes %>% 
  rename(Subject.accession = V1) %>% 
  mutate(Subject.accession = str_remove(Subject.accession, "\\.\\d+"))
```

Filter relevant hits:

-   Minimum alignment length: 190 nt
-   Remove species in ban list;
-   Remove hits from non-target genes:
-   Keep hits from relevant biological groups (bony fishes, whales and dolphins)

```{r, eval = FALSE}
# Filer valid hits
filtered_hits <- all_hits %>%
  filter(Alignment.length >= 190,
        !Scientific.name %in% ban_list,
         Subject.accession %in% target_genes$Subject.accession,
         Subject.blast.name %in% c("bony fishes","whales & dolphins"))
```

After filtration, we have multiple hits for each ASV. To obtain the best hit, we select the hits with highest bit score and percentage identity:

```{r, eval = FALSE}
# Obtain top hits and Remove environmental samples hits before summarizing
top_hits <- filtered_hits %>%
    # Remove environmental sample rows
  filter(!grepl("environmental sample", Scientific.name, ignore.case = TRUE)) %>%
  # Normalize to first two words for species-level matching
  mutate(Scientific.name = sub("^([A-Za-z]+\\s+[A-Za-z]+).*", "\\1", Scientific.name)) %>%
  group_by(Query.accession) %>%
  filter(Bit.Score == max(Bit.Score)) %>%
  filter(Percentage.of.identical.matches == max(Percentage.of.identical.matches)) %>% 
  ungroup()
```

It is possible to obtain multiple hits with the same scores, but different species. Usually, within the same genus or within the same family. We call these situations *ties*.

To solve ties we need the following functions:

-   "./R/check_ties.R" *identifies ties that need to be solved*
-   "./R/assign_LCA.R" *breaks the tie by identifying the last common ancestor - LCA*

Additionally, we also need a reference list of genera within Families that we want to break ties.

**Note:**

-   For species tied within the same genus, we can always assign LCA - it is the genus sp.
-   For species tied between different genera, but within the same family, we can break the tie, if we have a reference. For now, we have references for the families **Pleuronectidae**, **Ziphiidae**, **Salmonidae**, **Mugilidae** and **Delphinidae**.
-   For ties outside the above mentioned situations, we cannot establish LCA and we are not confident on the best hit, therefore, we assigned them as **Uncertain**.
-   Note: if genera from different families are tied, we assign to **Uncertain**.

Load the functions into your R session:

```{r, eval = FALSE}
source("./R/check_ties.R")
source("./R/assign_LCA.R")
```

Load the reference list for families **Pleuronectidae**, **Ziphiidae** and **Delphinidae**.

```{r, eval = FALSE}
# Add reference for families
delphinidae_family <- read.table("delphinidae_family.txt"); names(delphinidae_family) <- "Genus"
pleuronectidae_family <- read.table("pleuronectidae_family.txt"); names(pleuronectidae_family) <- "Genus"
ziphiidae_family <- read.table("ziphiidae_family.txt"); names(ziphiidae_family) <- "Genus"
salmonidae_family <- read.table("Salmonidae_family.txt"); names(salmonidae_family) <- "Genus"
mugilidae_family <- read.table("Mugilidae_family.txt"); names(mugilidae_family) <- "Genus"
```

Get best hits, without ties:

```{r, eval = FALSE}
# best hits, with LCA
taxonomic_assignments <- top_hits %>%
  group_by(Query.accession) %>% 
  nest() %>% 
  mutate(LCA = map(.x = data, 
                   .f = ~assign_LCA(.x))) %>% 
  mutate(taxa = map(.x = data, .f = ~unique(.x$Scientific.name))) %>% 
  mutate(isTie = map(.x = taxa, .f = ~ifelse(length(unique(.x)) == 1, FALSE, TRUE))) %>% 
  unnest(c(LCA, data, isTie)) %>% 
  group_by(Query.accession) %>% 
  slice_head(n = 1) %>% 
  mutate(FinalAssignment = ifelse(isTRUE(isTie), LCA, Scientific.name)) %>% 
  select(Query.accession,
         Scientific.name, 
         LCA, FinalAssignment, 
         Bit.Score, evalue, Alignment.length,
         Percentage.of.identical.matches,
         Number.of.identical.matches,
         Number.of.mismatches)

# view results in your R session
View(taxonomic_assignments)

# Save final assignments into memory
write.csv(taxonomic_assignments, "taxonomic_assignments.csv")
```

## Add taxonomic assignments to ASV abundance table

First, we need to transform the blast results in a data frame compatible with the abundance table. Then, we can merge them based on ASV ID.

```{r, eval = FALSE}
# transform blast results to compatible format
ASV_ncbi <- taxonomic_assignments %>% 
  select(ASV = Query.accession, Scientific.name) %>% 
  filter(!is.na(Scientific.name)) %>%  # Remove unassigned ASVs
  left_join(ASVs.df, by = "ASV") # ASVs.df was made in the DADA2 section

# Create abundance table (long format)
abundance_table_long <- ASV_table %>% # ASV_table was made in DADA2 section
  prepare_tidy_data(sample_names = row.names(ASV_table), samples_in = "rows") %>% 
  rename(Sequence = Taxa_id) %>% 
  left_join(ASVs.df, by = "Sequence") %>% 
  left_join(ASV_ncbi, by = "ASV")
```

## Extras

After the abundance table is ready, we can add additional steps, like the removal of ASVs below 0.1% relative abundance.

```{r, eval = FALSE}
# Filter local low abundance (< 0.01%)
total_reads <- abundance_table_long %>%
  group_by(Sample) %>%
  summarise(total = sum(Abundance, na.rm = TRUE))

abundance_table_long_filtered <- abundance_table_long %>%
  left_join(total_reads, by = "Sample") %>%
  mutate(freq = Abundance / total * 100,
         Abundance = ifelse(freq < 0.01, 0, Abundance)) %>%
  select(-total, -freq)
```

**Identify ASVs from control samples and filter them out**

```{r, eval = FALSE}
# Create control map, connecting samples to their controls
sample_control_map <- list(
  "M1-1-16S_S1_L001_R1_001" = c("CE1-16S_S1_L001_R1_001", "CF1-1-16S_S1_L001_R1_001"),
  "M1-2-16S_S1_L001_R1_001" = c("CE1-16S_S1_L001_R1_001", "CF1-2-16S_S1_L001_R1_001"),
  "M1-3-16S_S1_L001_R1_001" = c("CE1-16S_S1_L001_R1_001", "CF1-3-16S_S1_L001_R1_001"),
  "M2-1-16S_S1_L001_R1_001" = c("CE2-16S_S1_L001_R1_001", "CF2-1-16S_S1_L001_R1_001"),
  "M2-1-NZY-16S_S1_L001_R1_001" = c("CE2-16S_S1_L001_R1_001", "CF2-1-16S_S1_L001_R1_001"),
  "M2-2-16S_S1_L001_R1_001" = c("CE2-16S_S1_L001_R1_001", "CF2-2-16S_S1_L001_R1_001"),
  "M2-2-NZY-16S_S1_L001_R1_001" = c("CE2-16S_S1_L001_R1_001", "CF2-2-16S_S1_L001_R1_001"),
  "M2-3-16S_S1_L001_R1_001" = c("CE2-16S_S1_L001_R1_001", "CF2-3-16S_S1_L001_R1_001"),
  "M2-3-NZY-16S_S1_L001_R1_001" = c("CE2-16S_S1_L001_R1_001", "CF2-3-16S_S1_L001_R1_001"),
  "M3-1-16S_S1_L001_R1_001" = c("CE3-16S_S1_L001_R1_001", "CF2-1-16S_S1_L001_R1_001"),
  "M3-2-16S_S1_L001_R1_001" = c("CE3-16S_S1_L001_R1_001", "CF2-2-16S_S1_L001_R1_001"),
  "M3-3-16S_S1_L001_R1_001" = c("CE3-16S_S1_L001_R1_001", "CF2-3-16S_S1_L001_R1_001")
)

# Identify ASVs present in control samples
asvs_in_controls <- abundance_table_long_filtered %>%
  filter(Sample %in% unlist(sample_control_map),
         Abundance > 0) %>%
  distinct(Sample, ASV)

# Remove those ASVs from their corresponding environmental samples
for (sample_name in names(sample_control_map)) {
  controls <- sample_control_map[[sample_name]]
  
  contaminant_asvs <- asvs_in_controls %>%
    filter(Sample %in% controls) %>%
    pull(ASV) %>%
    unique()
  
  abundance_table_long_filtered <- abundance_table_long_filtered %>%
    mutate(Abundance = ifelse(Sample == sample_name & ASV %in% contaminant_asvs, 0, Abundance))
}

# Save final long-format abundance table
write.csv(abundance_table_long_filtered, file = "abundance_table_long_eDNA.csv", row.names = FALSE)

# Create wide-format abundance table
abundance_table_wide <- abundance_table_long_filtered %>% 
  pivot_wider(names_from = Sample, values_from = Abundance)

# Save wide-format table
write.csv(abundance_table_wide, file = "abundance_table_wide_eDNA.csv", row.names = FALSE)
```

# Verify results

We provide some examples of data analyses below.

## Rarefaction curves

```{r, eval = FALSE}
# rarefaction curve
library(vegan)
library(dplyr)

# remove unnecessary columns
ASV_matrix.1 <- abundance_table_wide %>% 
  select(-Sequence, -Scientific.name) 

#
asc_col <- ASV_matrix.1$ASV
ASV_matrix.1$ASV <- NULL
rownames(ASV_matrix.1) <- asc_col

#  
ASV_matrix <- ASV_matrix.1 %>% t()
# rarefaction curve
rarecurve(ASV_matrix, 
          step = 500, 
          xlab = "Sequencing depth",
          ylab = "Number of ASVs")
```

## Example of quick diversity analysis

```{r, eval = FALSE}
#
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vegan)
library(scales)

# Load abundance table in wide format
df <- read_csv("abundance_table_wide_eDNA.csv")

# Filter to keep only target taxa (non-NA Scientific.name)
df <- df %>% filter(!is.na(Scientific.name))

# Define and remove unwanted columns
cols_to_remove <- c(
  "Sequence.x", "ASV", "Sequence.y",
  "BPCR1NZY-16S_S1_L001_R1_001", "BPCRP-16S_S1_L001_R1_001",
  "CE1-16S_S1_L001_R1_001", "CE2-16S_S1_L001_R1_001", "CE3-16S_S1_L001_R1_001",
  "CF1-1-16S_S1_L001_R1_001", "CF1-2-16S_S1_L001_R1_001", "CF1-3-16S_S1_L001_R1_001",
  "CF2-1-16S_S1_L001_R1_001", "CF2-2-16S_S1_L001_R1_001", "CF2-3-16S_S1_L001_R1_001"
)
df_clean <- df %>% select(-any_of(cols_to_remove))

# Create ASV matrix
# Ensure unique rownames (from first column, presumably ASV IDs)
rownames(df_clean) <- make.unique(as.character(df_clean[[1]]))

# Extract abundance matrix (drop first column which was used as rownames)
asv_matrix <- df_clean[, -1]

# Clean sample names by removing suffix
colnames(asv_matrix) <- gsub("-16S_S1_L001_R1_001", "", colnames(asv_matrix))

# Transpose: samples as rows, ASVs as columns
asv_matrix_t <- t(asv_matrix)

# Remove empty samples (rows with sum 0)
asv_matrix_t <- asv_matrix_t[rowSums(asv_matrix_t) > 0, ]

# Clean sample names again from filtered matrix rownames (just to be sure)
rownames(asv_matrix_t) <- gsub("-16S_S1_L001_R1_001", "", rownames(asv_matrix_t))

# Alpha Diversity per Sample (Observed + Shannon)
# Calculate diversity metrics per sample
alpha_df <- data.frame(
  Sample = rownames(asv_matrix_t),
  Observed = rowSums(asv_matrix_t > 0),
  Shannon = diversity(asv_matrix_t, index = "shannon")
)

# Observed Richness (Boxplot only)
ggplot(alpha_df, aes(y = Observed)) +
  geom_boxplot(fill = "steelblue", alpha = 0.8) +
  labs(title = "Observed ASVs per Sample", x = "", y = "Observed Richness") +
  theme_minimal()

# Shannon Index (Boxplot only)
ggplot(alpha_df, aes(y = Shannon)) +
  geom_boxplot(fill = "darkgreen", alpha = 0.8) +
  labs(title = "Shannon Diversity per Sample", x = "", y = "Shannon Index") +
  theme_minimal()

# Beta Diversity (Bray-Curtis)
# Calculate Bray-Curtis dissimilarity
bray_dist <- vegdist(asv_matrix_t, method = "bray")

# Perform NMDS (k=2 dimensions)
set.seed(42); nmds_res <- metaMDS(bray_dist, k = 2, trymax = 100)

# Extract NMDS points
nmds_df <- as.data.frame(nmds_res$points)
colnames(nmds_df) <- c("NMDS1", "NMDS2")

# Add sample names
nmds_df$Sample <- rownames(asv_matrix_t)

# Print NMDS stress value
cat("NMDS stress:", round(nmds_res$stress, 4), "\n")

# Plot NMDS
ggplot(nmds_df, aes(x = NMDS1, y = NMDS2, label = Sample)) +
  geom_point(size = 3, color = "darkorange") +
  geom_text(vjust = -0.5, size = 3) +
  labs(title = paste("Beta Diversity (Bray-Curtis NMDS), Stress =", round(nmds_res$stress, 3)),
       x = "NMDS1", y = "NMDS2") +
  theme_minimal()

# Calculate Jaccard dissimilarity (presence/absence)
jaccard_dist <- vegdist(asv_matrix_t, method = "jaccard", binary = TRUE)

# NMDS on Jaccard
set.seed(42); nmds_jaccard <- metaMDS(jaccard_dist, k = 2, trymax = 100)

# Prepare dataframe
nmds_jaccard_df <- as.data.frame(nmds_jaccard$points)
colnames(nmds_jaccard_df) <- c("NMDS1", "NMDS2")
nmds_jaccard_df$Sample <- rownames(nmds_jaccard_df)

# Plot Jaccard NMDS
ggplot(nmds_jaccard_df, aes(x = NMDS1, y = NMDS2, label = Sample)) +
  geom_point(size = 3, color = "steelblue") +
  geom_text(vjust = -0.5, size = 3) +
  labs(title = paste("Beta Diversity (Jaccard NMDS), Stress =", round(nmds_jaccard$stress, 3)),
       x = "NMDS1", y = "NMDS2") +
  theme_minimal()

# Relative Abundance Plot
#Pivot data to long format
asv_long <- df_clean %>%
  pivot_longer(cols = -Scientific.name, names_to = "Sample", values_to = "Abundance")

#Clean sample names
asv_long$Sample <- gsub("-16S_S1_L001_R1_001", "", asv_long$Sample)

#Remove zero abundances
asv_long <- asv_long %>% filter(Abundance > 0)

#Calculate relative abundance per sample
asv_rel_abund <- asv_long %>%
  group_by(Sample) %>%
  mutate(RelAbund = Abundance / sum(Abundance)) %>%
  ungroup()

#Use all species directly as taxon labels
asv_rel_abund <- asv_rel_abund %>%
  mutate(Taxon = Scientific.name)

#Aggregate relative abundances by Sample and Taxon
plot_data <- asv_rel_abund %>%
  group_by(Sample, Taxon) %>%
  summarise(RelAbund = sum(RelAbund), .groups = "drop")

#Define custom color palette (20 colors, unnamed)
custom_colors <- c(
  "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd",
  "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
  "#a6cee3", "#1b9e77", "#d95f02", "#7570b3", "#e7298a",
  "#66a61e", "#e6ab02", "#a6761d", "#666666", "#f781bf"
)

#Plot stacked bar chart
ggplot(plot_data, aes(x = Sample, y = RelAbund, fill = Taxon)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Relative Abundance by Taxon",
    x = "Sample",
    y = "Relative Abundance"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 8),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = custom_colors) +
  guides(fill = guide_legend(title = "Taxon"))
```

# References

-   Andrews, S. (2010). FastQC: A Quality Control Tool for High Throughput Sequence Data [Online]. Available online at: <http://www.bioinformatics.babraham.ac.uk/projects/fastqc/>

-   Ewels P, Magnusson M, Lundin S, Käller M. MultiQC: summarize analysis results for multiple tools and samples in a single report. Bioinformatics. 2016 Oct 1;32(19):3047-8. doi: 10.1093/bioinformatics/btw354. Epub 2016 Jun 16. PMID: 27312411; PMCID: PMC5039924.

-   Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJ, Holmes SP. DADA2: High-resolution sample inference from Illumina amplicon data. Nat Methods. 2016 Jul;13(7):581-3. doi: 10.1038/nmeth.3869. Epub 2016 May 23. PMID: 27214047; PMCID: PMC4927377.

-   Camacho, C., Coulouris, G., Avagyan, V., Ma, N., Papadopoulos, J., Bealer, K., and Madden, T.L. 2009. BLAST+: architecture and applications. BMC Bioinformatics, 10, 421.

-   Altschul, S.F., Gish, W., Miller, W., Myers, E.W. and Lipman, D.J., 1990. Basic local alignment search tool. Journal of Molecular Biology, 215(3), pp.403-410.

-   Benson, D. A., Cavanaugh, M., Clark, K., Karsch-Mizrachi, I., Lipman, D. J., Ostell, J., & Sayers, E. W. (2013). GenBank. Nucleic acids research, 41(D1), D36-D42.
